Transferability of clinical prediction models for early trauma care

### 2019-07-29 ManuscriptMarkdown.Rmd
Update ManuscriptMarkdown.Rmd

---
title: "Transferability of clinical prediction models for early trauma care"
author: 
- Martin Henriksson
- Dell Saulnier
- Johanna Berg
- Martin Gerdin Wärnberg
date: "July 18, 2019"
documentclass: "article"
output:
  word_document:
    reference_docx: transferstyle.docx
header-includes:
   - \usepackage{hyperref}
bibliography: library.bib
csl: bmcemerg.csl
---
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)

@@

# Introduction
Trauma results in approximately 3,8 million deaths annually across the globe, 
and road traffic injuries are among the top 10 leading causes of death [@RN1; @RN21]. 
Each year 9% of the world’s deaths are because of trauma, and
predictions indicate that this number is likely to increase [@RN42].

In a typical, high resource setting, the initial trauma management is
performed on the scene by Emergency Medical Services
(EMS). Patient data and vital signs are transferred to the receiving hospital. 
This information is then evaluated using a system to determine
the level of trauma, prepare adequate resources [@RN43] and dictate if a full
or limited trauma team is activated [@RN44].

Systems to determine the level of trauma can be based on clinical prediction
models. Models differ in quality and characteristics, but generally perform
well predicting survival [@RN45]. Many of these models are developed in a single
standardized setting such as a major trauma centre but are then implemented in 
different settings, such as non-trauma centres.

@@

In trauma, misclassification is often referred to as mistriage. Triage refers to
the classification of severity as minor or major trauma. Mistriage can be
subdivided into overtriage; incorrect classification of a patient with minor
trauma as major trauma, or undertriage; incorrect classification of a patient
with major trauma as minor trauma. Both are detrimental in terms of patient care
and distribution of resources.

Thus, the effect of model transfers between contexts within the setting of a
single health care system, as well as the effect of such transfers on mistriage
have not been studied and represent substantial knowledge gaps. Therefore, the
aim of this study is to assess how transfers of clinical prediction models for
early trauma care between different contexts within a single health system
affect mistriage rates.

# Materials and methods
## Study design
A registry-based cohort study was conducted. Sweden has a nationally encompassing
trauma register called SweTrau. SweTrau data was used to create
clinical prediction models, which were transferred between different contexts to
study the effects on mistriage. The study and analysis plans were made
publically available before the research was undertaken [@RN59].

@@

The trauma care organization in Sweden is usually clearly defined in major
metropolitan areas: trauma patients are transported directly to a predesignated
hospital with specific trauma competency. In large, more rural parts of the
country, trauma patients are instead transported to the nearest hospital for
stabilization. Once stabilized, they may then be transported to hospitals with
more advanced trauma care capacity [@RN34]. Most hospitals in Sweden use some
form of system to categorize the level of trauma. These systems are usually
based on patient vital signs and injury mechanism, as reported by the EMS, or as
registered in the emergency department [@RN90].

Today 95,5% (52 of 55) of Swedish hospitals record trauma cases in SweTrau. 
Currently the registry includes 55 000 cases, and the board encourages its 
use in both academic research and local quality improvement initiatives [@RN28].

## Participants 
The eligibility criteria were patients registered in SweTrau between 2011 and 2016, 
age 15 or above. This age was decided as the study aims to study adult trauma and not
paediatric trauma which differs in physiology, triage and initial care
[@RN46]. Also, the Swedish guidelines for trauma activation define children as age <15
[@RN90]. Patient age was obtained from the SweTrau register.

## Variables
### Model predictors
The clinical prediction models used the predictors systolic blood pressure
(SBP), respiratory rate (RR) and Glasgow coma scale (GCS), all on arrival to
hospital. The rationale for these three predictors is that they are
part of many established clinical prediction models for early trauma care, such
as the Revised Trauma Score [@RN35].

@@

### Participant characteristics
To describe the patient cohort we reported age, sex, Injury Severity Score (ISS)
and New ISS (NISS).

@@

## Data sources and measurements
Model predictors, outcome, participant characteristics and study outcome as
outlined above were all obtained from SweTrau. The method of measurement for the
model predictors is not specified in the registry entries. In Swedish emergency rooms,
patient parameters are usually obtained by a registered nurse or assistant nurse
and are assumed to be accurate.  Whether the patient is dead or alive 30 days
after trauma is manually entered into the registry locally by each respective
hospital. NISS and ISS are calculated by hospital personal
based on the patient’s injuries.

## Bias
Data analysis was conducted according to a prearranged analysis plan. The
analysis plan and statistical analysis code was finalised using simulated
data. These efforts were taken to avoid confirmation bias. The analysis plan was
reviewed by an experienced statistician and programmer prior to implementation. 
Neither outcome nor variable were blinded when conducting data analysis, which 
made a structured objective approach important.

## Study size
All patients matching eligibility criteria listed above. Four data sets were
used to study the transfer of clinical prediction models, each data set
representing a different setting. Data set, and sample size considerations are
outlined below.

## Quantitative variables
SBP and RR was modelled using restricted cubic splines with four knots whenever
possible placed at equally spaced percentiles, and GCS as a continuous linear
term. When describing the participant characteristics quantitative variables
were presented as continuous. ISS was presented as dichotomous using ISS > 15 as
the cutoff.

## Statistical methods
### Data sets
The complete SweTrau cohort was split into four sets of data, each representing a 
plausible clinical setting for model transfer. These data sets were further divided 
into samples, this process is outlined below.

### High and low volume centres
Based on number of patients received and registred in SweTrau for each centre, two 
samples were derived from this data set. High volume centres were those with in the 
top quartile of number of patients registred. The rest were low volume centres.

### Metropolitan and non-metropolitan centres
This data set was also split into two samples. The metropolitan sample consisted
of greater Stockholm, greater Gothenburg and greater Malmö, as defined by
statistics Sweden. The other sample was patients from non-metropolitan areas.

### Multi and single centre data
In this data set multiple samples were created. Each centre with large enough
sample size (outlined below) to develop and validate a model constituted their own 
sample. The multi-centre sample consisted of the combined data from all centres.

### Individual centres
This data set was also split into multiple samples. Each centre with large
enough sample size to develop and validate a model constituted its own
sample. This data set was removed in the final version of the study due to a
lack of centres with a large enough sample size.

### Development and validation sample
Each data set included at least two samples. The samples were then split into
two subsamples using a temporal split based on the date of trauma. The earlier
subsample was the development sample, and the later subsample the validation
sample. The development sample contained 70 events (events being patients who
died within 30 days of the trauma) and all non-events (non-events being patient
survival 30 days past the trauma) during the same time. The rationale for
including 70 events is that at least 10 events per free parameter were needed in
the logistic regression to obtain stable coefficient estimates [@RN30]. The
validation sample instead contained 100 events and at least 100 non-events,
which was suggested as the minimum number by Vergouwe in 2005 for external
validation samples [@RN37]. See figure 1 for example using the high- and low
volume centre data set. The minimum sample size of development and
validation samples were 140 and 200 respectively. We performed analyses only on
data sets for which all samples included at least the minimum sample size.

![](fig1.jpg)  
**Fig. 1: High and low volume centre data set.** Illustrates creation of development 
and validation sample in the high and low volume data set. Initial split based on
number of patients. Temporal split made using date of traumatic event.

### Sequence of analysis
The programming language R was used for all analyses [@RN31], and all code has
been made publically available [@RN59]. Analyses were performed in the sequence
of model development, model validation and finally model comparison. These steps
were repeated in each data set. Below we used the transfer of a model from a
high-volume sample to a low volume sample as an example to describe the complete
procedure.

### Model development
In the model development step a clinical prediction model was developed in the
high-volume centre development sample. The model was developed using logistic
regression. The dependent variable was all cause mortality within 30 days of 
trauma and independent variables SBP, RR, and GCS modelled as previously described. 
To avoid overfitting the model a bootstrap procedure was used to estimate a linear 
shrinkage factor that was then applied to the model coefficients [@RN61]. The shrunk 
model was then used to estimate the probability of all cause 30-day mortality in the 
development sample. A gridsearch was performed across estimated probabilities in 
the development sample to identify the cut-off that optimised overtriage keeping 
undertriage at less than 5% [@RN39]. This cut-off was then used to classify patients 
as major or minor trauma.

### Model validation
In the model validation step, the model performance was assessed in the
high-volume centre validation sample and in the low volume centre validation
data. First the model was used to estimate the probability of all cause 30-day
mortality in each sample. Then the probability cutoff identified in the
development sample was applied to the validation samples, patients were
classified as major or minor trauma, and model performance was estimated.

### Model comparison
Finally, in the model comparison step, the difference in model performance
between the high and low volume centre validation samples was
calculated. Empirical bootstrap was used to estimate 95% confidence intervals
(CI) around performance and differences in performance estimates. Both bootstrap
procedures used 1000 bootstrap samples drawn with replacement of the same size
as the original samples. The three steps of model development, model validation,
and model comparison were repeated in all four sets of data.

@@

### Missing data
We used multiple imputation using chained equations (mice), to handle missing data [@RN41]. 
The number of imputations created
for each data set was equal to the highest percentage of missing data in that
data set. Quantitative variables were imputed using predictive mean matching and
qualitative variables were imputed using logistic regression. SBP and RR was
transformed as restricted cubic splines before imputation. All analyses outlined 
above were then conducted separately in
each imputed dataset. We present the combined results as the median point
estimate across imputations along with an empirical bootstrap of the 25th and
75th percentiles across imputations, i.e. the lower bound of the presented
interval is the lower bound of a 95% CI of the 25th percentile and the upper
bound is the upper bound of a 95% CI of the 75th percentile. This combined CI
was referred to as an Uncertainty Interval (UI) and was used to express the
added uncertainty associated with the imputation procedure to handle missing
data, as such it is more conservative than a standard 95% CI.

## Ethical considerations
This study is approved by the regional ethics review board in Stockholm,
Sweden. Ethical review numbers are 2015/426-31 and 2016/461-32. All patients
included recived letters detainling SweTrau inclusion, and the possibility of
their data being used for scientific purposes. It was assumed that the patients
who consented to research want to contribute to this improvement of care, and as
thus by performing this study the authors hope to honour their wishes.

# Results
We analysed data from `r total.number.of.patients ` trauma patients to
investigate the effects of transfer on clinical prediction models (Table 1). The
total number of missing observations across all variables was `r total.number.of.na ` 
in the entire study cohort. The sample with the highest
percentage of missing observations was the `r highest.percent.missing.name `
sample with `r highest.percent.missing.number `% incomplete observations. The
variable with the highest number of missing values was `r name.variable.na `,
with `r number.variable.na ` missing values, which constituted `r round(percent.variable.na) `
% of the total values for this variable. The percentages 
of missing values for the other model predictors, GCS and SBP was `r round(gcs.na$percentage)`% 
and `r round(sbp.na$percentage)`% respectively.

@@

## Development and validation
During model development, the number of imputations used for each data set was
`r high.low.imp`, `r metro.non.imp` and `r multi.single.imp` for the high and
low volume data set, the metropolitan and non-metropolitan data set and the
multi and single centre data set respectively.

Table 2 shows model validation performance. The model with the lowest validation
mistriage rate was the `r colnames(lowest.val.mistriage)` model with a median
mistriage rate of `r lowest.val.mistriage`. Performance in terms of
undertriage and overtriage rate for the same model was `r val.undertriage`, and
`r val.overtriage` respectively. Worst validation performance i.e. highest
mistriage rate was found in the `r colnames(highest.val.mistriage)` sample, with
a model median mistriage rate of `r highest.val.mistriage`.

>>>>

---
title: "Transferability of clinical prediction models for early trauma care"
author: 
- Martin Henriksson
- Dell Saulnier
- Johanna Berg
- Martin Gerdin Wärnberg
date: "July 29, 2019"
documentclass: "article"
output:
  word_document:
    reference_docx: transferstyle.docx
header-includes:
   - \usepackage{hyperref}
bibliography: library.bib
csl: bmcemerg.csl
---
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)

@@

# Introduction
Trauma, defined as an injury to living tissue by an extrinsic agent [@RN63], 
results in approximately 3,8 million deaths annually across the globe [@RN1; @RN21]. 
Each year 9% of the world’s deaths are because of trauma, with the leading causes being 
road traffic accidents, suicide and homicide. Predictions indicate the incidence of all 
these causes are likely to increase by 2030 [@RN42].

In a typical, high-resource setting, the initial management of trauma is
performed on the scene by Emergency Medical Services
(EMS). Patient data and vital signs are transferred to the receiving hospital. 
This information is then evaluated using a system to determine
the level of trauma, prepare adequate resources [@RN43] and dictate if a full
or limited trauma team is activated [@RN44].

Systems that determine the level of trauma can be based on clinical prediction
models. Models differ in quality and characteristics, but generally perform
well at predicting survival [@RN45]. Many models are developed in a single,
standardized setting, such as a major trauma centre but are then implemented in 
different settings, such as non-trauma centres.

@@

In trauma, misclassification is often referred to as mistriage. Triage refers to
the classification of severity as minor or major trauma. Mistriage can be
subdivided into overtriage, the incorrect classification of a patient with minor
trauma as major trauma, or undertriage, the incorrect classification of a patient
with major trauma as minor trauma. Mistriage can ultimatly lead to decreased patient 
survival, but is also detrimental in terms of patient care and distribution of resources.

Thus, the effect of model transfers between care contexts within a single healthcare system, 
as well as the effect of such transfers on mistriage,have not been studied and represent 
substantial knowledge gaps. Therefore, the aim of this study is to assess how transfers 
of clinical prediction models for early (on arrival to hospital) trauma 
care between different care contexts within a single health system affect mistriage rates.

# Materials and methods
## Study design
A registry-based cohort study was conducted. Sweden has a nationally encompassing
trauma register called SweTrau. SweTrau data was used to create
clinical prediction models, which were transferred between different care contexts to
study the effects on mistriage. The study and analysis plans were made
publicly available before the research was undertaken [@RN59].

@@

The trauma care in Sweden is usually clearly defined in major
metropolitan areas: trauma patients are transported directly to a predesignated
hospital with specific trauma competency. In more rural parts of the
country, trauma patients are instead transported to the nearest hospital for
stabilization. Once stabilized, they may then be transported to hospitals with
more advanced trauma care capacity [@RN34]. Most hospitals in Sweden use some
form of system to categorize the level of trauma, usually
based on patient vital signs and injury mechanism, as reported by the EMS, or as
registered in the emergency department [@RN90].

Today 52 of the 55 Swedish hospitals record trauma cases in SweTrau. 
Currently, the registry includes 55 000 cases, and its use is encouraged in 
both academic research and local quality improvement initiatives [@RN28].

## Participants 
The eligibility criteria were adult patients aged 15 years or above registered in SweTrau 
between 2011 and 2016. The study aimed to assess adult trauma and not
paediatric trauma which differs in physiology, triage and initial care
[@RN46]. The Swedish guidelines for trauma activation define children as age <15
[@RN90].

## Variables
### Participant characteristics
To describe the patient cohort we reported age, sex, Injury Severity Score (ISS)
and New ISS (NISS).

### Model predictors
The clinical prediction models used the predictors systolic blood pressure
(SBP), respiratory rate (RR) and Glasgow coma scale (GCS), all on arrival to
hospital. The rationale for these three predictors is that they are
part of multiple established clinical prediction models for early trauma care, such
as the Revised Trauma Score [@RN35].

@@

@@

## Data sources and measurements
Model predictors, outcome, participant characteristics and study outcome as
outlined above were all obtained from SweTrau. The method of measurement for the
model predictors is not specified in the registry entries. In Swedish emergency rooms,
patient parameters are usually obtained by a registered nurse or assistant nurse
and are assumed to be accurate.  Whether the patient is dead or alive 30 days
after trauma is manually entered into the registry by each respective
hospital. NISS and ISS are calculated by hospital personnel
based on the patient’s injuries.

## Bias
Data analysis was conducted according to a prearranged analysis plan. The
analysis plan and statistical analysis code was finalised using simulated
data. These efforts were taken to avoid confirmation bias. The analysis plan was
reviewed by an experienced statistician and programmer prior to implementation. 
Neither the outcome nor variables were blinded when during analysis, which 
made a structured approach important, to ensure objectivity.

## Study size
All patients matching eligibility criteria were included. Four data sets were
used to study the transfer of clinical prediction models, each data set
representing a different care context. The data sets, and sample size considerations 
are outlined below.

## Quantitative variables
GCS was modelled as a continuous linear term, and SBP and RR were modelled using 
restricted cubic splines with four knots whenever possible, placed at equally spaced 
percentiles. When describing the participant characteristics, quantitative variables 
were presented as continuous. ISS was presented as dichotomous using ISS > 15 as the cutoff.

## Statistical methods
### Data sets
The complete SweTrau cohort was split into three overlapping sets of data, each representing a 
plausible care context for model transfer. These data sets were further divided 
into samples, this process is outlined below.

### Data set 1: High and low volume centres
Two samples were derived from the SweTrau cohort, based on the number of patients. 
High volume centres were those within the top quartile of number of patients received. 
The rest were low volume centres.

### Data set 2: Metropolitan and non-metropolitan centres
This data set was also split into two samples: a metropolitan sample consisting of 
patients from greater Stockholm, greater Gothenburg and greater Malmö, as defined by 
statistics Sweden, and a second sample of patients from non-metropolitan areas.

### Data set 3: Multi- and single centre data
In this data set, two samples were created. One centre with a large enough 
sample size to develop and validate a model was its own sample. The multicentre 
sample consisted of the combined data from all centres.

A fourth dataset was initially created, consisting of individual centres that 
had a large enough sample size to develop and validate a model. However, this 
data set was removed before analysis due to a lack of centres 
with a large enough sample size.

### Development and validation sample
The samples in each data set were split into
two subsamples using a temporal split based on the date of trauma. The earlier
subsample was the development sample, and the later subsample the validation
sample. The development sample contained 70 events (events being patients who
died within 30 days of the trauma) and all non-events (non-events being patient
survival 30 days past the trauma) during the same time. The rationale for
including 70 events is that at least 10 events per free parameter were needed in
the logistic regression to obtain stable coefficient estimates [@RN30]. The
validation sample contained 100 events and at least 100 non-events,
which was suggested as the minimum number by Vergouwe in 2005 for external
validation samples [@RN37]. Figure 1 shows the development of the samples using 
the high- and low volume centre data set as an example. The minimum sample size of development and
validation samples were 140 and 200 respectively. We performed analyses only on
data sets for which all samples included at least the minimum sample size.

![](fig1.jpg)  
**Fig. 1: Sample development in the high and low volume centre data set.** Illustrates 
creation of development and validation sample in the high and low volume data set. 
Initial split based on number of patients. Temporal split made using date of traumatic event.

### Sequence of analysis
The programming language R was used for all analyses [@RN31], and all code has
been made publicly available [@RN59]. Analyses were performed in the sequence
of model development, model validation and finally model comparison. These steps
were repeated in each data set. Below, we use the transfer of a model from the
high volume centre sample to a low volume centre sample as an example to describe 
the complete procedure.

### Model development
In the model development step, a clinical prediction model was developed in the
high volume centre development sample. The model was developed using logistic
regression. The dependent variable was all-cause mortality within 30 days of 
trauma and the independent variables SBP, RR, and GCS were modelled as previously described. 
To avoid overfitting the model, a bootstrap procedure was used to estimate a linear 
shrinkage factor that was then applied to the model coefficients [@RN61]. The shrunk 
model was used to estimate the probability of all-cause 30-day mortality in the 
development sample. A gridsearch was performed across estimated probabilities in 
the development sample to identify the cut-off that optimised overtriage keeping 
undertriage at less than 5% [@RN39]. This cut-off was then used to classify patients 
as major or minor trauma.

### Model validation
In the model validation step, the model performance was assessed in the
high volume centre validation sample and in the low volume centre validation
sample. First, the model was used to estimate the probability of all-cause 30-day
mortality in each sample. Then the probability cut-off identified in the
development sample was applied to the validation samples, patients were
classified as major or minor trauma, and model performance was estimated.

### Model comparison
Finally, the difference in model performance
between the high and low volume centre validation samples was
calculated. Empirical bootstrap was used to estimate 95% confidence intervals
(CI) around performance and differences in performance estimates. Both bootstrap
procedures used 1 000 bootstrap samples drawn with replacement of the same size
as the original samples. The three steps of model development, model validation,
and model comparison were repeated in all four sets of data.

@@

### Missing data
We used multiple imputation using chained equations (mice) to handle missing data [@RN41]. 
The number of imputations created
for each data set was equal to the highest percentage of missing data in that
data set. Quantitative variables were imputed using predictive mean matching and
qualitative variables were imputed using logistic regression. SBP and RR was
transformed as restricted cubic splines before imputation. All analyses outlined 
above were then conducted separately in
each imputed dataset. We present the combined results as the median point
estimate across imputations along with an empirical bootstrap of the 25th and
75th percentiles across imputations, i.e. the lower bound of the presented
interval is the lower bound of a 95% CI of the 25th percentile and the upper
bound is the upper bound of a 95% CI of the 75th percentile. This combined CI
was referred to as an Uncertainty Interval (UI) and was used to express the
added uncertainty associated with the imputation procedure to handle missing
data; as such, it is more conservative than a standard 95% CI.

## Ethical considerations
The study was approved by the regional ethics review board in Stockholm, Sweden. 
Ethical review numbers are 2015/426-31 and 2016/461-32. All patients in SweTrau 
had received letters detailing their inclusion in the database, and the possibility 
of their data being used for scientific purposes. It was assumed that the patients 
who consented to the use of their SweTrau data for research want to contribute to 
the improvement of care, and thus by performing this study the authors hope to 
honour their wishes.

# Results
We analysed data from `r total.number.of.patients ` trauma patients to
investigate the effects of transfer on clinical prediction models (Table 1). The
total number of missing observations across all variables was `r total.number.of.na ` 
in the entire study cohort. The sample with the highest
percentage of missing observations was the `r highest.percent.missing.name `
sample with `r highest.percent.missing.number `% incomplete observations. The
variable with the highest number of missing values was `r name.variable.na `,
with `r number.variable.na ` missing values, or `r round(percent.variable.na)`% of 
the total values for this variable. The percentages of missing values for the other 
model predictors, was `r round(gcs.na$percentage)`% for GCS and `r round(sbp.na$percentage)`% 
for SBP.

@@

## Development and validation
During model development, the number of imputations used for each data set was
`r high.low.imp`, `r metro.non.imp` and `r multi.single.imp` for the high and
low volume data set, the metropolitan and non-metropolitan data set, and the
multi- and single centre data set, respectively.

Table 2 shows model validation performance. The model with the lowest validation
mistriage rate was the `r colnames(lowest.val.mistriage)` model with a median
mistriage rate of `r lowest.val.mistriage`. Performance in terms of
undertriage and overtriage rate for the same model was `r val.undertriage`, and
`r val.overtriage` respectively. Worst validation performance (i.e. highest
mistriage rate) was found in the `r colnames(highest.val.mistriage)` sample, with
a model median mistriage rate of `r highest.val.mistriage`.
